Draft final report\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-11-07 draftfinalreport.Rnw
revised text NF comments

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks of measles after measles introductions from abroad. In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand. Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007. We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of naive people at the District Health Board (DHB) and national level. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
\item Estimates of the proportion of the currently naive New Zealand population (11\%) requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item Peak overall travel rates, and thus presumably risk from measles importation, is in December. However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this 2013--2014 outbreak, as analysed from data until June, is well above one. This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
\item The mean wage losses per measles case is estimated to be \$852
\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
\item The mean management cost per case was \$1,765
\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>>$1).
\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
\end{itemize}

@@

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sec:}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age. Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

@@

\section{Additional risk analyses}
\label{sub:risk_analyses}

@@

%%%%%%%%%%%%%%%%%%
\subsection{Risk of measles infection in New Zealand analyses methods}
\label{sub:name}

We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand were required to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago. We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of risk of measles infection.

@@

The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories. For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}). 
<<echo=FALSE,include=FALSE,results=hide>>=
## regression analyses - measles
# setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

@@

ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
                 data = testtable , FUN=sum)
@

\begin{table}[hbtp]
\footnotesize
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(ttyr, file="",table.env=FALSE,rowname=NULL)
@
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
\label{table:case}
\end{table}

<<echo=FALSE,include=FALSE,results=hide>>=
### denominator data
# setwd("~/Massey_2014/measles/data")
denom<-read.csv("NZDep2006Denominators.csv",header=T)
head(denom)
denom<-denom[,-c(9)]
summary(denom)
dim(denom)
denom<-denom[,1:18]
row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
sum(row.has.na)
denomt <- denom[!row.has.na,]
head(denomt)
denomt<-denomt[denomt$Sex_code==99,]
denomt$Eth_Level<-as.factor(denomt$Eth_Level)
denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
dm<-cbind(denomt[,c(3,6,9:18)])
head(dm)
dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
dm$Age_Label<-dage
colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
head(dm)
datadm<-dm[dm$Age=="<6",]
dataA<-(rbind(datadm,datadm)/2)
dataA<-round(dataA[,3:12])
dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
ddm<-dm[!dm$Age=="<6",]
dm<-rbind(dataA,ddm)
dm$Ethnicity <- factor(dm$Ethnicity)
deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
dm$Ethnicity<-deth
head(dm)

@@

tp$cases <- 0
cases <- matrix(0, length(tp$merge),1)
rownames(cases) <- tp$merge
cases[ttyr$merge,] <- ttyr$Cases
tp$cases <- cases
PerCapita<-(tp$cases/tp$Popn)
PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita,4))
colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
@

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table~\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
par(mar=c(5,6,4,2)+0.1)
par(cex.axis=2)
plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="Category",cex.lab=2)
legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
       col=c(1:4,6),pch=16,bty="n",cex=1.8)
@
\end{center}
\caption{Per capita rates of measles infections for 2007--2014 broken down by ethnicity (Table~\ref{table:percap} for details)}
\label{fig:percap}
\end{figure}

\begin{table}[hbtp]
\footnotesize
\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@
\label{table:percap}
\end{table}

@@

\subsection{Vaccination history of measles infection and population immunity estimation methods}

@@

The distribution of the meases cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
@
\end{center}
\caption{Distribution of measles cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

@@

The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
\par
\vspace{5mm} %5mm vertical space
\begin{table}
<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
library(xtable)
resan<-xtable(anova(model3,test="F"))
print(resan,floating = FALSE)
@
\caption{Significance of different predictor variables for measles risk factors from 2007-2014}
\label{table:anova}
\end{table}
\vspace{5mm} %5mm vertical space
A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.

@@

People of Pacific origin are also over-represented as measles cases ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:}). Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.

@@

The majority of the naive among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}). However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}). Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}). The distribution of the numbers of naive and the total naive populations per DHB, assuming national immunisation and immunity rates are representative, are given in Figures~\ref{fig:Northland},\ref{fig:Waitemata},\ref{fig:Auckland},\ref{fig:Counties_Manukau},\ref{fig:Waikato},\ref{fig:Lakes},\ref{fig:BayofPlenty},\ref{fig:Tairawhiti},\ref{fig:Taranaki},\ref{fig:HawkesBay},\ref{fig:Whanganui},\ref{fig:Midcentral},\ref{fig:Hutt},\ref{fig:CapitalandCoast},\ref{fig:Wairarapa},\ref{fig:NelsonMarlborough},\ref{fig:WestCoast},\ref{fig:Canterbury},\ref{fig:SouthCanterbury}, and \ref{fig:Southern}.

@@

%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
library(xtable)
totdpdf<-read.csv("mapdata_tot.csv",header=T)
# output top countries
colnames(totdpdf)[1]<-c("country")
toptot <- totdpdf[order(-totdpdf$immigration),] 
immigration<-toptot[1:30,c(1,4)]
topincidence <- totdpdf[order(-totdpdf$incidence),] 
incidence<-topincidence[1:30,c(1,2)]
incidence[,2]<-round(incidence[,2],0)
topvaccine <- totdpdf[order(totdpdf$cover),] 
vaccinecover<-topvaccine[1:30,c(1,3)]
toprisk <- totdpdf[order(-totdpdf$risk),] 
risk<-(toprisk[1:30,c(1,5)])
risk[,2]<-round(risk[,2],0)
@

@@

Analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:imtravel12}. However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}. Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.Vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12}\ref{fig:incidence12}. Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand. These data are shown in Table \ref{table:risk12}. The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}. The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12}\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12}\ref{fig:imrisk12}

>>>>

New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks of measles following measles introductions from abroad. In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region. We identified additional analyses that may help understand risk of infection in New Zealand. Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007. We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, considering pre- and post-vaccination scenarios, based on the numbers of naive people at the District Health Board (DHB) and national level. We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels. Our key findings were:

\begin{itemize}
\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
\item The regression analyses suggest that age is a particularly strong risk factor for measles. However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis. In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age. People of European ethnicity, however, make up the majority of cases.
\item Of the 11\% of the population of New Zealand that are currently considered to be immunologically naive, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
\item Peak overall travel rates, and thus presumably risk from measles importation, is in December. However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and the current 2013--2014 outbreak, as analysed from data until June, is well above one. This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
\item The mean wage losses per measles case is estimated to be \$852
\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
\item The mean management cost per case was \$1,765
\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1).
\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
\end{itemize}

@@

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above). Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age. Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

@@

Thus in this report we include regression analyses that help understand risk factors for being measles cases in New zealand (Section~\ref{sub:risk_analyses}, descriptions of the New Zealand population in terms of vaccination coverage and immunity (Section~\ref{sub:immunestat}), and use those data to model the likely measles outbreak sizes in each Distric Health Board and the country (Section~\ref{sec:epidemic_modelling}. We estimate the costs of previous and current measles outbreaks (Section~\ref{sub:cost}) and uses these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch up vaccination programmes (Section~\ref{sub:cost_benefit}). This report includes further developments of the analyses provided in previous interim reports.

\section{Risk analyses}
\label{sub:risk_analyses}

@@

%%%%%%%%%%%%%%%%%%
\subsection{Risk of measles infection in New Zealand analyses methods}

We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to adjust for confounding due to a lack of independence among risk factors. Specifically $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand were required to adjust for confounding and test whether interactions among covariates provide additional information on risk in demographically--defined subgroups over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago. We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of measles infection risk.

@@

The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories. For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}). 
<<echo=FALSE,include=FALSE,results=hide>>=
## regression analyses - measles
# setwd("~/Massey 2014/DHayman_20140627")
# read data
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

@@

ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
                 data = testtable , FUN=sum)
ttyr <- ttyr[ttyr$Ethnicity=="None",]
@

\begin{table}[hbtp]
\footnotesize
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(ttyr, file="",table.env=FALSE,rowname=NULL)
@
\caption{Absolute number of measles cases in specific age and socio-economic deprivation categories from 2007-2014 with no ethnicity recorded}
\label{table:none}
\end{table}

<<echo=FALSE,include=FALSE,results=hide>>=
### denominator data
# setwd("~/Massey_2014/measles/data")
data<-read.csv("DHayman_20140627.csv",header=T)
vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
names(data)
data$RptYear<-as.factor(data$RptYear)
data$SurvWeek<-as.factor(data$SurvWeek)
data$NZDep01<-as.factor(data$NZDep01)
data$NZDep06<-as.factor(data$NZDep06)
data$NZDep13<-as.factor(data$NZDep13)
#data$AgeInYears<-as.factor(data$AgeInYears)
data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)

ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
                 data = testtable , FUN=sum)
denom<-read.csv("NZDep2006Denominators.csv",header=T)
head(denom)
denom<-denom[,-c(9)]
summary(denom)
dim(denom)
denom<-denom[,1:18]
row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
sum(row.has.na)
denomt <- denom[!row.has.na,]
head(denomt)
denomt<-denomt[denomt$Sex_code==99,]
denomt$Eth_Level<-as.factor(denomt$Eth_Level)
denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
dm<-cbind(denomt[,c(3,6,9:18)])
head(dm)
dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
dm$Age_Label<-dage
colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
head(dm)
datadm<-dm[dm$Age=="<6",]
dataA<-(rbind(datadm,datadm)/2)
dataA<-round(dataA[,3:12])
dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
ddm<-dm[!dm$Age=="<6",]
dm<-rbind(dataA,ddm)
dm$Ethnicity <- factor(dm$Ethnicity)
deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
dm$Ethnicity<-deth
head(dm)

@@

tp$cases <- 0
cases <- matrix(0, length(tp$merge),1)
rownames(cases) <- tp$merge
cases[ttyr$merge,] <- ttyr$Cases
tp$cases <- cases
PerCapita<-(tp$cases/tp$Popn)
PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita*10000,4))
colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
@

The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}. Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses. However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category. Along with small population sizes (Table~\ref{table:percap}), there are several issues with estimating the denominator data for this group (University of Otago, personal communication). Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities. This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).

\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 15,height=11>>=
par(mar=c(12,6,4,2)+0.1)
par(cex.axis=2)
plot(tp$cases/tp$Popn*10000,main="",ylab="Per Capita Rate Per 10,000",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
     xlab="",cex.lab=2,xaxt="n")
axis(side = 1, at = 1:50, labels=rep(c("0-2yr, 1-5","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),las=2,cex.axis=1.5)
mtext(side=1,line=10,text="Age and NZDep",cex=2)
legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
       col=c(1:4,6),pch=16,bty="n",cex=1.8)
par(mar = c(5,4,4,2) + 0.1)
@
\end{center}
\caption{Per capita rates of measles infections for specific age and New Zealand deprivation indices (NZDep) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
\label{fig:percap}
\end{figure}

\begin{table}[hbtp]
\footnotesize
\caption{Numbers of measles cases, population sizes and per capita rates per 10,000 of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@
\label{table:percap}
\end{table}

@@

\subsection{Vaccination history of measles infection and population immunity estimation methods}
\label{sub:immunestat}

@@

The distribution of the meases cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
\begin{figure}
\begin{center}
<<fig=TRUE,echo=FALSE,width = 11>>=
par(cex.axis=2)
par(mar=c(5,6,4,2)+0.1)
hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
@
\end{center}
\caption{Distribution of measles cases per category used in the final regression model}
\label{fig:histcasecat}
\end{figure}

@@

The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
\par
\vspace{5mm} %5mm vertical space
\begin{table}
<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
library(xtable)
resan<-xtable(anova(model3,test="F"))
print(resan,floating = FALSE)
@
\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
\label{table:anova}
\end{table}
\vspace{5mm} %5mm vertical space
A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.

@@

The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001). Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes. In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected. The very young (<2 years old) are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}). This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}). Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.

@@

The majority of the naive among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}). However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the 3--5 year old age group to greater levels than in older young people  (Figure~\ref{fig:naive}). Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}). The distribution of the numbers of naive and the total naive populations per DHB, assuming national immunisation and immunity rates are representative, are given in Figures~\ref{fig:Northland}, \ref{fig:Waitemata}, \ref{fig:Auckland}, \ref{fig:Counties_Manukau}, \ref{fig:Waikato}, \ref{fig:Lakes}, \ref{fig:BayofPlenty}, \ref{fig:Tairawhiti}, \ref{fig:Taranaki}, \ref{fig:HawkesBay}, \ref{fig:Whanganui}, \ref{fig:Midcentral}, \ref{fig:Hutt}, \ref{fig:CapitalandCoast}, \ref{fig:Wairarapa}, \ref{fig:NelsonMarlborough}, \ref{fig:WestCoast}, \ref{fig:Canterbury}, \ref{fig:SouthCanterbury}, and \ref{fig:Southern}.

@@

%%%%%%%%%%%%%%%%%%
<<echo=FALSE>>=
library(xtable)
totdpdf<-read.csv("mapdata_tot.csv",header=T)
# output top countries
colnames(totdpdf)[1]<-c("country")
toptot <- totdpdf[order(-totdpdf$immigration),] 
immigration<-toptot[1:30,c(1,4)]
topincidence <- totdpdf[order(-totdpdf$incidence),] 
incidence<-topincidence[1:30,c(1,2)]
incidence[,2]<-round(incidence[,2],0)
topvaccine <- totdpdf[order(totdpdf$cover),] 
vaccinecover<-topvaccine[1:30,c(1,3)]
toprisk <- totdpdf[order(-totdpdf$risk),] 
risk<-(toprisk[1:30,c(1,5)])
risk[,2]<-round(risk[,2],0)
toprisk2<-(toprisk[1:10,c(1,2,3,4)])
toprisk2[,2]<-round(toprisk2[,2],0)
@

@@

\begin{table}
\caption{Measles and travel data from 2012 for the top 10 countries identified as high risk for measles importation. incidence: measles incidence per million; cover: percent vaccination cover; immigration: the number of travellers entering New Zealand from the country}
<<results=tex,echo=FALSE>>=
# This chunk uses "results=tex"
library(Hmisc)
latex(toprisk2, file="", table.env=FALSE,rowname=NULL)
@
\label{table:toprisk}
\end{table}

Analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}. However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}. Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}. The

Vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12}\ref{fig:incidence12}. Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand. These data are shown in Table \ref{table:risk12}. The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}. The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12}\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12}\ref{fig:imrisk12}
